Bird beak's elaboration and exploration

### 2021-06-07 Analysis/03-elaboration_exploration_analyses.Rmd
Update 03-elaboration_exploration_analyses.Rmd
Minore edits and typos

All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
Some analyses in `beer` do take some time to run so we advice to not drive or operate heavy machinery while using it.
Don't use it while driving.
The following code snippets are used to illustrate the implementation of the functions.
See <!-- §§§run_replicates_script --> for running the actual analyses.

@@

To analyse the exploration and elaboration aspects of both clades and tips, we first selected a random 1000 covariance matrices from the combined MCMCglmm posterior variance-covariance matrices.

@@

We will use these covariance matrices as the basis of the phylogenetic main axes of variation.
These axes correspond to the main axes of variation in the $N_DIMENSIONS$ trait space at different levels:

@@

These major axes where calculated as the longest distance within the 95% confidence interval hyper-ellipse from the variance-covariance matrix (i.e. the major axes of the 95% CI ellipse of matrix).
In practice, the coordinates of these axes end points were calculated directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigen values and the variance-covariance matrix's eigen vectors.

## Major axes calculation

For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and of radius being composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:

@@

(in other words $O_{n}$ is the matrix representing each edge point of the unit hyper sphere of $n$ dimensions and of radius $1$);

And $O'_{n}$ the scaled matrix hyper sphere to the right confidence interval size using the $\chi^2$ distribution:

@@

We can get $M_{n}$ matrix containing all the edges' coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scale 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:

@@

Where $M_{1,n}$ is the major axes of the 0.95 CI hyper ellipse fitting the variance-covariance matrix, $M_{2,n}$ the second second axes (minor axes in 2D), etc...

Finally, the centred the matrix $M_{1,m}$ on the centroid of each clade.

@@

We can then compare these axes either between each other or relative to the tips positions in the trait space using elaboration and exploration scores:

@@

From these major axes, we can use linear algebra in $n$ dimensions to estimate clades or tips' exploration and elaboration scores.
These scores are a direct interpretation of @endler2005's elaboration and innovation (here exploration) patterns (Fig 1. in @endler2005):
Given a major axis of evolutionary "trajectory" (in the sense of evolutionary trend, not an evolutionary direction), on can calculate the elaboration score as the projection value of any vector defined by a clade or a tip on the major axis and the exploration (or "innovation" in @endler2005) as that vector's rejection.

@@

Figure: in a space with 5 elements: A, B, C, D, E (in grey); where D and E represent the major axes (e.g. the major phylogenetic effect axes from the MCMCglmm), we can rotate and rescale each element so that D and E become the unit vector of length 1 (the black letters D and E) and get the exploration elaboration scores for either the elements (e.g. element A's projection in blue and its rejection in orange) or from any other axes (e.g. the major phylogenetic effect axes from the MCMCglmm from a specific clade).

@@

 1. In the trait space, define $\vec{B}$ as the base vector (typically $\vec{B}$ is defined as the pair of coordinates set from the major axis described above).
 2. Centre the trait space on the origin of $\vec{B}$ so that the first set of coordinates of $\vec{B}$ are 0.
 3. Scale the trait space to the norm of $\vec{B}$ so that the norm of $\vec{B}$ is now 1.
 4. Rotate the trait space using the rotation matrix $R_{\vec{B}}$ to satisfy the linear transformation $\vec{B} \arrow \vec{\hat{\imath}}$ (with $\vec{\hat{\imath}}$ being the first unit vector of the trait space - typically the x axes unit vector). 
 5. Project/reject every element in the trait on $\vec{B}$ (that is now $\vec{\hat{\imath}}$). In practice, the first coordinate (x) of each element is now it's projection onto $\vec{B}$.

Following the ways to:
 1. estimate the variance-covariance matrices of the phylogenetic effects in trait space (02-MCMCglmm_mini_chains.Rmd);
 2. extracting the major axes of variation of these VCV matrices for the different levels in the phylogeny;
 3. and calculating the elaboration and exploration scores for elements in the trait space based on these major axes;

@@

To run the analyses per clade we compared the major axes for each clade to the major axes for the overall phylogenetic effect (e.g. plovers' VCV vs. all birds' VCV, etc.).

To do so, for randomly selected posterior variance-covariance matrix we defined $\vec{a}$ as the major axes for the clade of interest and projected it onto $\vec{B}$ as the major overall phylogenetic effect (see above).
For each pair of $\vec{a}$ and $\vec{B}$, we translated the origin of $\vec{a}$ onto the origin of $\vec{B}$ and measured the projection (elaboration), rejection (exploration) and angle (degree of exploration).

@@

Similarly to the analyses described above, we ran the analyses per species but here we directly compared each species position in the space to the same vector $\vec{B}$ described above and reported the results per groups.

@@

Note that here only the tips positions projected on the global phylogenetic axes.

Finally, we also looked at the exploration elaboration profiles within each clades by replacing for each clade, the projection of the elements onto $\vec{B}$ to the projection onto their respective clade major axes $\vec{a}$.

>>>>

All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
Although `beer` is intentionally designed to be portable and shareable, we advise workers to use it with moderation and tailor its consumption to their specific research needs.
Some analyses in `beer` do take some time to run so we advise users not to operate heavy machinery while using it.
Don't use it while driving.
The following code snippets are used to illustrate the implementation of the functions.
See <!-- §§§run_replicates_script --> for running the actual analyses.

@@

To analyse the exploration and elaboration aspects of both clades and tips, we first selected 1000 random covariance matrices from the combined MCMCglmm posterior variance-covariance matrices.

@@

We use these covariance matrices as the basis of the phylogenetic main axes of variation.
These axes correspond to the main axes of variation in the $N_DIMENSIONS$ trait space at different phylogenetic levels:

@@

These major axes were calculated as the longest distance within the 95% confidence interval hyper-ellipse from the variance-covariance matrix (i.e. the major axes of the 95% CI ellipse of matrix). % NC: A small figure here would help
In practice, the coordinates of these axes end points were calculated directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigenvalues and the variance-covariance matrix's eigenvectors.

## Major axes calculation
% NC: Mathematically seems sound, but I'm no expert. But very hard to parse for anyone who hasn't been looking at this in detail. Think fo the poor palaeontologists!

For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and a radius composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:

@@

(in other words $O_{n}$ is the matrix representing each edge point of the unit hypersphere of $n$ dimensions and of radius $1$);

And $O'_{n}$ is the scaled matrix hypersphere to the 95\% confidence interval size using the $\chi^2$ distribution:

@@

We can get $M_{n}$ matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scale 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:

@@

Where $M_{1,n}$ is the major axis of the 0.95 CI hyper-ellipse fitting the variance-covariance matrix, $M_{2,n}$ the second axis (minor axis in 2D), etc. etc.

Finally, we centred the matrix $M_{1,m}$ on the centroid of each clade.

@@

We can then compare these axes either with each other or relative to the tip positions in the trait space using elaboration and exploration scores as described below.

@@

From the major axes calculated above, we can use linear algebra in $n$ dimensions to estimate exploration and elaboration scores for clades or tips of the tree.
These scores are a direct interpretation of @endler2005's elaboration and innovation (here exploration) patterns (Fig 1. in @endler2005):
Given a major axis of evolutionary "trajectory" (in the sense of evolutionary trend, not an evolutionary direction), we can calculate the elaboration score as the projection value of any vector defined by a clade or a tip on the major axis and the exploration (or "innovation" in @endler2005) as that vector's rejection.

@@

Figure: in a space with five elements: A, B, C, D, E (in grey); where D and E represent the major axes (e.g. the major phylogenetic effect axes from the MCMCglmm), we can rotate and rescale each element so that D and E become the unit vector of length 1 (the black letters D and E) and get the exploration and elaboration scores for either the elements (e.g. element A's projection in blue and its rejection in orange) or from any other axes (e.g. the major phylogenetic effect axes from the MCMCglmm for a specific clade).

@@

 1. In the trait space, define $\vec{B}$ as the base vector (typically $\vec{B}$ is defined as the pair of coordinates set from the major axis described above).
 2. Centre the trait space on the origin of $\vec{B}$ so that the first set of coordinates of $\vec{B}$ are 0.
 3. Scale the trait space to the norm of $\vec{B}$ so that the norm of $\vec{B}$ is now 1.
 4. Rotate the trait space using the rotation matrix $R_{\vec{B}}$ to satisfy the linear transformation $\vec{B} \arrow \vec{\hat{\imath}}$ (with $\vec{\hat{\imath}}$ being the first unit vector of the trait space - typically the x axes unit vector). 
 5. Project/reject every element in the trait on $\vec{B}$ (that is now $\vec{\hat{\imath}}$). In practice, the first coordinate (x) of each element is now its projection onto $\vec{B}$.

Following the ways to: %NC: Not sure what you mean here
 1. estimate the variance-covariance matrices of the phylogenetic effects in trait space (02-MCMCglmm_mini_chains.Rmd);
 2. extracting the major axes of variation of these VCV matrices for the different levels in the phylogeny;
 3. and calculating the elaboration and exploration scores for elements in the trait space based on these major axes.

@@

To run the analyses for each clade we compared the major axes for each clade to the major axes for the overall phylogenetic effect (e.g. plovers' VCV vs. all birds' VCV, etc.).

To do so, for a randomly selected posterior variance-covariance matrix we defined $\vec{a}$ as the major axes for the clade of interest and projected it onto $\vec{B}$ as the major overall phylogenetic effect (see above).
For each pair of $\vec{a}$ and $\vec{B}$, we translated the origin of $\vec{a}$ onto the origin of $\vec{B}$ and measured the projection (elaboration), rejection (exploration) and angle (degree of exploration).

@@

Similarly to the analyses described above, we ran the analyses for each species but here we directly compared each species position in the space to the same vector $\vec{B}$ described above and reported the results for each groups.

@@

Note that here only the tip positions are projected on the global phylogenetic axes.

Finally, we also looked at the exploration elaboration profiles within each clade, by replacing the projection of the elements onto $\vec{B}$ for each clade to the projection onto their respective clade major axes $\vec{a}$.
